---
title: 'Reproducible Research: Course Project 2'
author: "Rhys Hewer"
date: "22 August 2018"
output:
  html_document:
    keep_md: yes
---

# Analysis of Storm Impact on Fatalities, Injuries and Damage Costs.

## Synopsis

The U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

This analysis focuses on analysing the most impactful storms and weather events with respect to fatalities, injuries and economic damage.

Particularly, the exercise is to answer the questions:

* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

The analysis will show tornadoes to be the most damaging to people and floods to have the greatest economic impact.

## Data Processing

First steps is to load the relevant libraries

```{r message=FALSE}
##load libraries
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(tidyr)
```

Then to download and read the data.

```{r message=FALSE, cache=TRUE}
##download & read data
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, destfile = "stormdata.csv.bz2")
data <- read_csv("stormdata.csv.bz2")
```

Let's narrow down the data to just the information that we will need for the analysis. I have chosen variables which relate to health and economics.

```{r echo=TRUE}
##Select just the data required to answer the questions
data <- data %>% select("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")

kable(head(data)) %>%
        kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```
  
Tidy data is:

* Each variable forms a column.
* Each observation forms a row.
* Each type of observational unit forms a table.
  
The data is not tidy. "FATALITIES" and "INJURIES" are descriptions of a type of health effect and should be values within a variable. "PROPDMG" and "CROPDMG" are descriptions of a type of damage and should be values within a variable.

Health damage and economic damage are different observational units.

As such, data to be tidied so that variable "effect" shows the type of health effect of the event (fatality, injury) accompanied by the number of the effect in "number", stored in one table. "type" shows the type of damage done (property/crops) accompanied by the cost of that damage in "cost" stored in a different table.

#### Health data processing

```{r echo=TRUE}
##Tidy health related data
health <- data %>% 
        gather(FATALITIES, INJURIES, key = effect, value = number) %>%
        select(EVTYPE, effect, number)

healthHarm <- health %>% group_by(EVTYPE, effect) %>% 
        summarise(total = sum(number)) 
```

#### Economic data processing

The values in "PROPDMGEXP" and "CROPDMGEXP" are multiplication factors for the "PROPDMG" and "CROPDMG" variables. Information sourced from here: https://rpubs.com/flyingdisc/PROPDMGEXP

As such, we need to convert letter codes to the multiplication factors and multiply out the damage variables to get to the true damage costs.

```{r echo=TRUE}
##Select economic data
econ <-data %>% select("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")

##Change letter code for mulitplication factor for PROPDMGEXP
econ$PROPDMGEXP <- econ$PROPDMGEXP %>% recode(
        "K" = 10^3,
        "M" = 10^6,
        "B" = 10^9,
        "m" = 10^6,
        "+" = 1,
        "0" = 10,
        "5" = 10,
        "6" = 10,
        "?" = 10,
        "4" = 10,
        "2" = 10,
        "3" = 10,
        "h" = 10^2,
        "7" = 10,
        "H" = 10^2,
        "-" = 0,
        "1" = 10,
        "8" = 10)

econ$PROPDMGEXP[is.na(econ$PROPDMGEXP)] <- 0

##Change letter code for mulitplication factor for CROPDMGEXP
econ$CROPDMGEXP <- econ$CROPDMGEXP %>% recode(
        "K" = 10^3,
        "M" = 10^6,
        "B" = 10^9,
        "m" = 10^6,
        "0" = 10,
        "?" = 10,
        "2" = 10,
        "k" = 10^3)

econ$CROPDMGEXP[is.na(econ$CROPDMGEXP)] <- 0

##create total columns for property and crop damage
econ <- econ %>% mutate(propCost = PROPDMG * PROPDMGEXP)
econ <- econ %>% mutate(cropCost = CROPDMG * CROPDMGEXP)

econ <- econ %>% select("EVTYPE", "propCost", "cropCost")
```

We still need to tidy the economic data as mentioned previously.

```{r echo=TRUE}
##Tidy economic data
econ <- econ %>% gather(propCost, cropCost, key = "type", value = "cost")
```

## Results  

#### Health Results

Let's look at the top 10 causes of fatalities

```{r echo=TRUE}
##fatality results
hhfat <- healthHarm %>%
        filter(effect == "FATALITIES") %>%
        arrange(desc(total)) %>%
        head(10)

##Plot fatalities
hf <- ggplot(hhfat, aes(reorder(EVTYPE, total), total)) + 
        geom_col(fill ="#5489dd") +
        coord_flip() +
        ylab("Fatalities") + 
        xlab("Event") +
        ggtitle("Fatalities per weather event")
print(hf)
```

Description: Top 10 weather event causes of fatalities.


Let's look at the top 10 causes of injuries

```{r echo=TRUE}
##Injury results
hhinj <- healthHarm %>% 
        filter(effect == "INJURIES") %>%
        arrange(desc(total)) %>%
        head(10)

##Plot Injuries
hi <- ggplot(hhinj, aes(reorder(EVTYPE, total), total)) + 
        geom_col(fill = "#7d8e39") + 
        coord_flip() +
        ylab("Injuries") + 
        xlab("Event") +
        ggtitle("Injuries per weather event")
print(hi)
```

Description: Top 10 weather event causes of injuries

#### Health conclusions

* Tornado is the biggest killer, with a fatality rate about 3 times as high as the number 2 cause.
* Tornado is also the most injurious, causing 10 times as many injuries as the number 2 cause.

The overall conclusion is that Tornadoes are the types of events that are most harmful to population health.  

#### Economic Results

Lets sum the crop and property costs for a total cost variable and look at the top 10.

```{r echo=TRUE}
##Economic totals
ecSum <- econ %>% group_by(EVTYPE, type) %>%
        summarise(cost = sum(cost)) 


ecTot <- ecSum %>% group_by(EVTYPE) %>%
        summarise(cost = sum(cost)) %>%
        arrange(desc(cost)) %>%
        head(10)

##convert to $billions
ecTot$cost <- ecTot$cost/(10^9)

##Plot economic costs
ec <- ggplot(ecTot, aes(reorder(EVTYPE, cost), cost)) + 
        geom_col(fill = "#855596") +
        coord_flip() +
        ylab("Cost ($Billion)") + 
        xlab("Event") +
        ggtitle("Total cost of damage per weather event (Top 10)")
print(ec)
```

Description: Top 10 most economically damaging weather events.

#### Economic conclusions

* Floods are the type of event to have the greatest economic consequences, about twice as costly as the number 2 cause.

